### 2020-04-28 manuscript/manuscript.tex
reknit after fixing path

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

Phylogenomic studies based on hundreds to thousands of genes have
greatly improved our understanding of evolutionary relationships among
major animal lineages. However, a considerable debate still remains
regarding the animal origin, with Ctenophora-sister and Porifera-sister
emerging as the two primary hypotheses, comprising one of the most
difficult to resolve nodes in animal tree of life. Here, we
systematically explore the mechanisms that potentially cause the
incongruent results by synthesizing data and results from all previous
phylogenomic analyses, and perform new analyses with these data in a
consistent manner to characterize the differences between studies. We
find that gene sampling varies across studies, indicating that different
regions of the genomes are sampled with little overlap among data
matrices. More importantly, we find Porifera-sister can only be
recovered with more constraints, including using site-heterogeneous CAT
model and inclusive of the most closely related outgroups in several
matrices, whereas Ctenophora-sister is recovered in all other cases. We
next quantify the support of phylogenetic signal over the two hypotheses
and find substantial differences in the distribution of phylogenetic
signal between models and outgroup choice (especially CAT models). Last,
we show that recoding can be a problematic method for addressing
compositional variation in phylogenetic analyses. Our results provide a
comprehensive overview of the current understanding of the first
diverging animal branch and why different studies yield different
answers. Rather than embracing one hypothesis over another, we hope our
efforts provides an integrative overview of the challenge and provides
direction for future studies to address difficult phylogenetic problems.

\hypertarget{main}{%
\subsection{Main}\label{main}}

Phylogenomic analyses with genome-scale data have the potential to
revolutionize our understanding of the tree of life. However, over the
past decade, there has been considerable debate about the position of
the root of the animal phylogeny, with Ctenophora-sister (comb jelly)
and Porifera-sister (sponges) (Fig. 1) emerging as the two primary
hypotheses. Historically, there was little debate about the root of the
animal tree of life and Porifera-sister was widely accepted though
rarely tested. By contrast, there has long been uncertainty about the
relationship of Ctenophora to other animals\textsuperscript{1}. An
understanding of the first diverging animal lineage is critical for
scientists to reconstruct events that occurred early in animal
evolution, especially for the evolution of complex structures such as
cell types, nervous systems, and developmental patterns. However, a wide
consensus of animal origin has not been reached despite multiple
efforts.

The first phylogenomic study to include ctenophores\textsuperscript{2}
suggested a new hypothesis, now referred to as Ctenophora-sister, that
ctenophores are our most distant animal relative. Since then many more
studies have been published and extensive reanalyzed, some supporting
Ctenophora-sister, some Porifera-sister, and some neither (Table 1). It
has become clear that this is a very difficult phylogenetic challenge,
and as the problem has become better characterized, it has become an
interesting test-case to phylogenetic biologists beyond those concerned
with this particular biological problem. Work has been hindered, though,
because it has been difficult to directly compare results across studies
and synthesize findings to understand the broader patterns of support.
Here we synthesize data and results from all previous phylogenomic
analyses that tested Ctenophora-sister and Porifera-sister, and
reanalyze these data using standardized methods, and perform new
analyses to characterize differences between studies. We hope that this
effort provides an integrative overview of the challenge and provides
direction for future studies. We also hope that the work we have
conducted here, including consolidating all the datasets in one place
with consistent formats and species names, will enhance the technical
value of this interesting question to methods-focused investigators that
look to develop methods to address difficult phylogenetic problems.

\includegraphics{figures/Figure_overview.png} \textbf{Fig. 1.} (A) The
Ctenophora-sister hypothesis posits that there is a clade (designated by
the orange node) that includes all animals except Ctenophora, and that
Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis
posits that there is a clade (designated by the green node) that
includes all animals except Porifera, and that Porifera is sister to
this clade. Testing these hypotheses requires evaluating the support for
each of these alternative nodes. (C) The animals and their outgroup
choice, showing the three progressively more inclusive clades
Choanimalia, Holozoa, and Opisthokonta (designated by the blue node).
(D) A ctenophore. (E) A sponge.

@@

Models of molecular evolution have several components that each consider
different aspects of the evolutionary process. The models that have been
used to model protein evolution in studies of the deep phylogeny have
largely differed according to three components: the exchangeability
matrix \(E\), the rate of evolution, and the state equilibrium
frequencies \(\Pi\). A detailed description of each component is in
Supplementary Results. Briefly, (1) the exchangeability matrix \(E\)
describes the rate at which one amino acid changes to another.
Exchangeability matrices have been used in the studies under
consideration here include: Poisson (or F81), WAG, LG, GTR, as they have
been used regularly in phylogenomic analyses; (2) While the
exchangeability matrix describes the relative rate of different changes
between amino acids, the actual rate can be further scaled, such as site
homogeneous rates and Gamma-rate heterogeneity; (3) the vector of
equilibrium frequencies \(\Pi\) describes the stationary frequency of
amino acids, such as empirical site-homogeneous, estimated
site-homogeneous, CAT site-heterogeneous\textsuperscript{3}, and C10-C60
models (C models).

Models can be assembled by selecting different options for all these
different components. The models that are applied in practice area
heavily influenced by engineering and computational costs, as well as
convention. For example, on the questions considered here, Poisson and
GTR exchangeability matrices have only been used in combination with CAT
site-heterogeneous models of equilibrium frequency. LG and WAG
exchangeability matrices have only been used with site homogeneous
estimates of equilibrium frequency. This is further confused by the
abbreviations that are used for models. Papers often discuss CAT and WAG
models as if they are exclusive, but these particular terms apply to
non-exclusive model components-- CAT refers to variation across sites
and WAG a particular exchangeability matrix. CAT is generally shorthand
for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium
frequency estimation. One could, though, run a WAG+CAT model, although
the computation time is much longer than Poisson+CAT model. To avoid
confusion on this point, we always specify the exchangeability matrix
first, followed by modifiers that describe the accommodation of
heterogeneity in equilibrium frequencies (\emph{e.g.}, CAT). If there
are no modifiers, then it is implied that site homogeneous models are
used. Moreover, the Gamma-rate heterogeinity is used in almost every
analysis.

\hypertarget{gene-sampling}{%
\subsubsection{Gene sampling}\label{gene-sampling}}

High-throughput sequencing is leading to exciting new data matrices for
phylogenomic analyses, potentially including hundreds or thousands of
genes to explore the root of animal phylogeny (Table 1). Although the
number of sampled genes might be different across studies, it is
reasonable to expect that a core set of orthologs should be present and
conserved across most animal genomes (\emph{e.g.}, animal BUSCO
genes\textsuperscript{4}). Surprisingly, gene composition and matrix
overlap has been rarely assessed across studies and most studies relied
on their own bioinformatic pipeline for determination of orthologs (for
details see Supplementary Results).

\hypertarget{outgroup-and-ingroup-taxon-sampling}{%
\subsubsection{Outgroup and ingroup taxon
sampling}\label{outgroup-and-ingroup-taxon-sampling}}

Outgroup selection (the taxa used to root the ingroup taxa) can
potentially affect the phylogenetic inference of animal evolution (Fig.
1C). Several studies have progressively removed more distantly related
outgroup taxa to test the effect of outgroup selection to ingroup
topology\textsuperscript{5,6}. Three progressively more inclusive clades
have often been investigated, including Choanomalia (Metazoa plus most
closely related Choanoflagellata), Holoza (Choanimalia plus more
distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has
been suggested the inclusion of more distant related (e.g.~Fungi)
outgroup taxa tend to recover Ctenophora-sister, whereas increasing
support of Porifera-sister often recovered when more closely related
outgroup taxa were used with site-heterogeneous CAT
models\textsuperscript{6}. In contrast of outgroup selection,
sensitivity to ingroup sampling has received less attention than
outgroup sampling. This may be because results have tended to be more
sensitive to outgroup sampling.

@@

Compositional heterogeneity (lineage-specific differences in amino acid
frequencies) was long thought has a major effect in resolving the animal
tree of life, especially for the deep animal
phylogeny\textsuperscript{7}. Variation in evolutionary rates across
taxa can lead to compositional heterogeneity that may lead to long
branch attraction of sponges or ctenophores\textsuperscript{8}. Feuda
\emph{et al}.\textsuperscript{9} has proposed to use data-recoding
method to reduce compositional heterogeneity in matrices. By discarding
information on which specific amino acid is found at each site in each
species and instead focusing on which groups those amino acids belong
to, they sought to reduce the impact of differences between species in
which particular amino acids within those groups are most frequent. By
reanalyzing Chang et al.~2015\textsuperscript{10} and Whelan et
al.~2015\textsuperscript{11}, they report that posterior predictive
analyses indicate recoded analyses have better model adequacy than
non-recoding analyses, and that Porifera-sister was strongly favored
under all recoding strategies using GTR-CAT models\textsuperscript{9}.
However, how the recoding methods impact phylogenetic inference has not
been well explored. Moreover, a recent simulation study has suggested
that non-recoding approaches significantly outperformed recoding
strategies\textsuperscript{12}. Thus, it is important to reevalute the
performance of the impact of data-recoding compared to non-recoding
methods in animal phylogeny.

@@

Here, we synthesize data and results from all previous phylogenomic
analyses that tested Ctenophora-sister and Porifera-sister, including 17
data matrices from 12 studies (Table 1).

\begin{longtable}[]{@{}lllrrrlll@{}}
\caption{Table 1. Overview of data matrices used in this
study}\tabularnewline
\toprule
Study & Matrix & Clade & Number of gene partitions & Number of sites &
Number of taxa & Infered sister lineage & Journal &
Citations\tabularnewline
\midrule
\endfirsthead
\toprule
Study & Matrix & Clade & Number of gene partitions & Number of sites &
Number of taxa & Infered sister lineage & Journal &
Citations\tabularnewline
\midrule
\endhead
Dunn2008 & Dunn2008 & Opisthokonta & 150 & 21152 & 64 &
Ctenophora-sister & Nature & NA\tabularnewline
Philippe2009 & Philippe2009 & Opisthokonta & 129 & 30257 & 55 &
Porifera-sister & Current Biology & NA\tabularnewline
Hejnol2009 & Hejnol2009 & Opisthokonta & 1487 & 270580 & 94 &
Ctenophora-sister & Proc. Biol. Sci. & NA\tabularnewline
NA & Hejnol2009\_subsampled & Opisthokonta & 844 & 55594 & 94 &
Ctenophora-sister & NA & NA\tabularnewline
Nosenko2013 & Nosenko2013\_nonribo\_9187 & Opisthokonta & 35 & 9187 & 63
& Ctenophora-sister & Molecular Phylogenetics and Evolution &
NA\tabularnewline
Nosenko2013 & Nosenko2013\_ribo\_14615 & Opisthokonta & 87 & 14615 & 71
& Porifera-sister & NA & NA\tabularnewline
Ryan2013 & Ryan2013\_est & Opisthokonta & 406 & 88384 & 61 &
Ctenophora-sister & Science & NA\tabularnewline
Moroz2014 & Moroz2014\_3d & Opisthokonta & 114 & 22772 & 46 &
Ctenophora-sister & Nature & NA\tabularnewline
Whelan2015 & Whelan2015\_D1 & Opisthokonta & 251 & 59733 & 76 &
Ctenophora-sister & PNAS & NA\tabularnewline
Whelan2015 & Whelan2015\_D10 & Opisthokonta & 210 & 59733 & 70 &
Ctenophora-sister & NA & NA\tabularnewline
Whelan2015 & Whelan2015\_D20 & Opisthokonta & 178 & 81008 & 65 &
Ctenophora-sister & NA & NA\tabularnewline
Borowiec2015 & Borowiec2015\_Best108 & Choaimalia & 108 & 41808 & 36 &
Ctenophora-sister & BMC Genomics & NA\tabularnewline
Borowiec2015 & Borowiec2015\_Total1080 & Choaimalia & 1080 & 384981 & 36
& Ctenophora-sister & NA & NA\tabularnewline
Pisani2015 & Ryan2013\_est & Opisthokonta & 406 & 88384 & 60 &
Porifera-sister & PNAS & NA\tabularnewline
NA & Whelan\_D6 & Opisthokonta & 115 & 33403 & 70 & Porifera-sister & NA
& NA\tabularnewline
Chang2015 & Chang2015 & Opisthokonta & 170 & 51940 & 77 &
Ctenophora-sister & PNAS & NA\tabularnewline
Simion2017 & Simion2017 & Opisthokonta & 1719 & 401632 & 97 &
Porifera-sister & Current Biology & NA\tabularnewline
NA & Simion2017\_subsampled & Opisthokonta & 1719 & 110602 & 97 &
Porifera-sister & NA & NA\tabularnewline
Feuda2017 & Chang2015 & Opisthokonta & 170 & 51940 & 77 &
Ctenophora-sister & Current Biology & NA\tabularnewline
NA & WhelanD20 & Opisthokonta & 178 & 81008 & 65 & Porifera-sister & NA
& NA\tabularnewline
Whelan2017 & Whelan2017b\_full & Holozoa & 212 & 68062 & 80 &
Ctenophora-sister & Nature Ecology and Evolution & NA\tabularnewline
Whelan2017 & Whelan2017b\_strict & Choaimalia & 59 & 49388 & 76 &
Ctenophora-sister & NA & NA\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{results-and-discussion}{%
\subsection{Results and Discussion}\label{results-and-discussion}}

@@

Taxa sampling plays a critical role in phylogenetic inference,
especially for the taxa whose phylogenetic position is poorly
understood. However, several data matrices have relatively skewed taxon
sampling, with more than 50\% taxa sampled in bilaterians compared to
other major animal lineages (Extended Data Fig. 1). These skewed data
matrices include two studies constructed from ESTs
(Dunn2008\textsuperscript{2} and Hejnol2009\textsuperscript{13}) and one
study from whole-genome data (Borrowiec2015\textsuperscript{14}).
Importantly, several earlier studies only include less than five
ctenophore species (Extended Data Fig. 1). The rest data matrices have
relatively balanced taxon composition, although the total number of
sampled taxon varies a lot. In general, increasing taxon sampling is
broadly accepted to improve phylogenetic inference and aid in the
placement of the earliest diverging animal lineage. Thus, it should be
noted that Whelan2017\textsuperscript{15} (has the most Ctenophore taxa)
and Simion2017\textsuperscript{16} (contains the most taxon sampling)
have much broader taxon sampling than other matrices.

\includegraphics{manuscript_files/figure-latex/taxon_rectangles-1.pdf}

\textbf{Fig. Extended Data Fig. 1.} Each of the primary matrices
considered here, color coded by taxon sampling. Horizontal size is
proportional to the number of genes sampled, vertical size to the number
of taxa sampled.

\hypertarget{matrix-gene-composition-and-overlap}{%
\subsubsection{Matrix gene composition and
overlap}\label{matrix-gene-composition-and-overlap}}

Gene sampling also plays an essential role in phylogenomic analyses, and
most often rely on the concatenation of sampled genes for deep animal
phylogeny. It has been suggested that the concatenation assumes that
phylogenetic signal from genes that do not share the species phylogeny
will be overwhelmed by the signal from the majority of genes whose
genealogy mirrors that of the species evolutionary
history\textsuperscript{17}. However, large phylogenomic data matrices
potentially contain genes with conflicting or weak phylogenetic signal
that can lead to incongruence phylogenomic results, especially when
analyzing deeper node represents ancient divergence in the
tree\textsuperscript{18}. Thus, it is important to evaluate the gene
composition and matrix overlap to explore level of congruence of gene
sampling across different matrices.

We first used the Metazoa BUSCO genes\textsuperscript{{\textbf{???}}} to
evaluate gene composition from all data matrices and the result is
unexpected (Extended Data Fig. 2A). The animal BUSCO dataset contains
978 single-copy orthologs (identified from 64 animal taxa) that are most
likely shared by most animal genomes, and are commonly used to assess
the completeness of genome or transcriptomic
assemblies\textsuperscript{4}. The BUSCO genes also are frequently used
as markers in many large-scale phylogenomic studies (e.g.,
yeasts\textsuperscript{19}, spiders\textsuperscript{20}). Thus, it is
reasonable to expect that these conserved genes should present at a
higher percentage than non-BUSCO genes in most matrices. However, we
found that gene partitions from all matrices comprised less than 50\% of
BUSCO genes, even with several data matrices contain less than 20
percent (Extended Data Fig. 2B). Moreover, we identified number of
ribosomal RNA genes in each data matrix (Extended Data Fig. 2C). We
found Philippe2009, Nosenko2013\_ribo\_14615 and Chang2015 matrices
possessed much higher content of ribosomal genes than other matrices
(Extended Data Fig. 2D). We also annotated each gene partition against
SwissProt\textsuperscript{21} and Gene Ontology databases, and results
are shown in Supplementary Table 1.

Previous analyses have mainly focused on outgroup selection and/or model
fits in different data matrices, and the matrix overlap across different
studies has been rarely investigated since all these studies rely on
their own method for orthology inference. Our initial intent of
comparing matrix overlap across matrices was to construct a new matrix
with shared taxon and gene sampling to test specific hypotheses about
differences in support. The hypothesis is that although the
bioinformatics pipeline used for OG determination is different, most
larger data matrices should share a core set of orthologs. However, our
results show that surprisingly little overlap across all the data
matrices from different studies (Fig. 2). Thus, these results strongly
suggested that gene sampling largely varies across studies, indicating
that different regions of the genomes are sampled with little overlap
among different data matrices. Moreover, the little overlap among
studies may potentially lead to incongruent phylogenetic results and
difficult to directly compare results across studies.

\includegraphics{manuscript_files/figure-latex/gene_composition,-1.pdf}

\textbf{Extended Data Fig. 2.} Annotation and representation of BUSCO
and ribosomal genes in each data matrix. (A). The number of partitions
with BUSCO annotations in each matrix, relative to the number of
partitions. (B). The percentage BUSCO annotations in each matrix. (C).
The number of partitions with annotations of ribosomal genes in each
matrix, relative to the number of partitions. (D). The percentage
ribosomal genes in each matrix.

\includegraphics{manuscript_files/figure-latex/matrix_overlap-1.pdf}

\textbf{Fig. 2.} Pairwise overlap between each of the primary matrices
considered here. Horizontal size is proportional to the number of genes
sampled, vertical size to the number of taxa sampled. The horizontal
intersection shows the proportions of shared genes, the vertical
intersection shows the proportions of shared taxa.

@@

\includegraphics{manuscript_files/figure-latex/support_published_analyses-1.pdf}
\textbf{Fig. 3.} A total of 163 analyses were transcribed from the
literature (Supplementary Table 2).

Here we summarized all the 163 previous phylogenetic analyses from 15
studies (Fig. 3 and Supplementary Table 2). Among these studies, the
main conclusions of five studies strongly in favor of Porifera-sister
and 10 in support of Ctenophora-sister (Table 1). Moreover, three
studies are secondary studies without any new data matrices.
Interestingly, we found that all analyses strongly support of
Ctenophora-sister once site-homogeneous models were used, no matter what
outgroup choice, phylogenetic inference and whether data-recoding are
used. The support for Porifera-sister increases with the exclusion of
more distantly related outgroups and with the use of site-heterogeneious
CAT (Poisson+CAT or GTR+CAT) models. It should be noted that many BI
analyses with CAT model are not well resolved (either not reach
convergence or do not provide significant support for the key node of
animal evolution) mainly due to computational limits of CAT models
employed in PhyloBayes. The strong support of Porifera-sister can only
be recovered from analyses using Poisson+CAT and recoding methods with
GTR+CAT models in several matrices. Although GTR+CAT model generally has
better model adequacy than Poisson+CAT or site-homogeneous models from
previous studies\textsuperscript{6,9}, Porifera-sister is not always
supported since several analyses strongly in favor Ctenophora-sister
even with only choanofalgellates are used as outgroups (Fig. 3). Thus,
our summary of previous phylogenetic analyses strongly indicated that
the trend of the support of Porifera-sister is generally increasing with
the use of more complex CAT models and more closely related outgroups,
whereas Ctenophora-sister is supported in all other cases.

\hypertarget{new-analyses-of-published-matrices}{%
\subsection{New analyses of published
matrices}\label{new-analyses-of-published-matrices}}

\includegraphics{manuscript_files/figure-latex/support_new_analyses-1.pdf}
\textbf{Fig. 4.} A total of 84 new phylgoenomic analyses were conducted
in this study (Supplementary Table 3).

>>>>

\hypertarget{summary-ideally-no-more-than-200-words-currently-251-words}{%
\subsection{Summary (ideally no more than 200 words, currently 251
words)}\label{summary-ideally-no-more-than-200-words-currently-251-words}}

XX This summary paragraph should be structured as follows: 2-3 sentences
of basic-level introduction to the field; a brief account of the
background and rationale of the work; a statement of the main
conclusions (introduced by the phrase `Here we show' or its equivalent);
and finally, 2-3 sentences putting the main findings into general
context so it is clear how the results described in the paper have moved
the field forwards

Phylogenomic studies based on hundreds to thousands of genes have
greatly improved our understanding of evolutionary relationships in the
animal tree of life. However, considerable debate still remains
regarding the deepest split in the animal tree of life, with
Ctenophora-sister and Porifera-sister emerging as the two primary
alternative hypotheses. This has been one of the most difficult to
resolve nodes in animal tree of life, and has major implications for our
understanding of the earliest events in animal evolution. Here we
systematically explore this question by synthesizing data and results
from all previous phylogenomic analyses and perform new analyses with
these data to characterize the differences between matrices under
consistent conditions. Our reanalyses are generally consistent with
previous results. We next quantify the support of phylogenetic signal
and conduct sensitive analyses in several representative data matrices.
We found that the Porifera-sister can only be recovered with several
combined factors, including inclusive of a more simplistic evolutionary
model, an unrealistic large number of substitution categories, more
closely related outgroups and certain gene samplings, whereas
Ctenophora-sister is supported in most other conditions. We also show
that recoding amino acid data into reduced state space, another approach
that has been suggested to address variation in amino acid frequencies,
is problematic. Thus, our results provide a comprehensive overview of
the current understanding of the first diverging animal branch and why
different studies yield different answers. This synthesis provides an
integrative overview of the challenge and provides direction for future
studies to address difficult phylogenetic problems.

\hypertarget{main-the-typical-length-of-an-article-with-3-4-modest-display-items-figures-and-tables-is-2000-2500-words-summary-paragraph-plus-body-text-currently-is-4500-words}{%
\subsection{Main (The typical length of an article with 3-4 modest
display items (figures and tables) is 2000-2500 words (summary paragraph
plus body text), currently is \textasciitilde{}4500
words\ldots{})}\label{main-the-typical-length-of-an-article-with-3-4-modest-display-items-figures-and-tables-is-2000-2500-words-summary-paragraph-plus-body-text-currently-is-4500-words}}

Phylogenomic analyses with genome-scale data have the potential to
revolutionize our understanding of the tree of life. However, over the
past decade, there has been considerable debate about the position of
the root of the animal phylogeny, with Ctenophora-sister (comb jelly)
and Porifera-sister (sponges) (Fig. 1) emerging as the two primary
hypotheses. Historically, there was little debate about the root of the
animal tree of life and Porifera-sister was widely accepted though
rarely tested. By contrast, there has long been uncertainty about the
relationship of Ctenophora to other animals\textsuperscript{1}. An
understanding of the placement of these lineages is critical for
scientists to reconstruct events that occurred early in animal
evolution, especially for the evolution of complex structures such as
cell types, nervous systems, and developmental patterns. However, a wide
consensus of animal origin has not been reached despite multiple
efforts.

The first phylogenomic study to include ctenophores\textsuperscript{2}
suggested a new hypothesis, now referred to as Ctenophora-sister, that
ctenophores are our most distant animal relative. Since then many more
studies have been published, some supporting Ctenophora-sister, some
Porifera-sister, and some neither (Table 1). It has become clear that
this is a very difficult phylogenetic challenge. As the problem has
become better characterized, it has become an interesting test-case to
phylogenetic biologists beyond those concerned with this particular
biological problem. Work has been hindered, though, because it has been
difficult to directly compare results across studies and synthesize
findings to understand the broader patterns of support. Here we
synthesize data and results from phylogenomic analyses that tested
Ctenophora-sister and Porifera-sister. We reanalyze these data using
standardized methods, and perform new analyses to characterize
differences between studies. This provides an integrative overview of
the challenges posed by this question and provides direction for future
studies. We also hope that the work we have conducted here, including
consolidating all the datasets in one place with consistent formats and
species names, will enhance the technical value of this interesting
question to methods-focused investigators that look to develop methods
to address difficult phylogenetic problems.

\includegraphics{figures/Figure_overview.png} \textbf{Fig.
XXFigure\_overview.} (A) The Ctenophora-sister hypothesis posits that
there is a clade (designated by the orange node) that includes all
animals except Ctenophora, and that Ctenophora is sister to this clade.
(B) The Porifera-sister hypothesis posits that there is a clade
(designated by the green node) that includes all animals except
Porifera, and that Porifera is sister to this clade. Testing these
hypotheses requires evaluating the support for each of these alternative
nodes. (C) The animals and their outgroup choice, showing the three
progressively more inclusive clades Choanimalia, Holozoa, and
Opisthokonta (designated by the blue node). (D) A ctenophore. (E) A
sponge.

@@

Models of molecular evolution have several components that each consider
different aspects of the evolutionary process. The models that have been
used to model protein evolution in studies of the deep phylogeny have
largely differed according to three components: the exchangeability
matrix \(E\) , the rate of evolution, and the amino acid equilibrium
frequencies \(\Pi\). The exchangeability matrix \(E\) describes the rate
at which one amino acid changes to another (\emph{e.g.} Poisson/F81, WAG
or LG, and GTR). The rate of exchange can be further scaled, for example
with a Gamma parameter that accommodates rate variation across sites.
Heterogeneous equilibrium approaches include CAT\textsuperscript{3},
which is implemented in the software PhyloBayes and has been widely
applied to questions of deep animal phylogeny. Models of molecular
evolution are assembled by selecting different options for all these
different components. The models that are applied in practice are
heavily influenced by computational costs, as well as convention. For
example, on the questions considered here, CAT site-heterogeneous models
of equilibrium frequency have only been used in combination with Poisson
and GTR exchangeability matrices. LG and WAG exchangeability matrices
have only been used with site homogeneous estimates of equilibrium
frequency. This is further confused by the abbreviations that are used
for models. Papers often discuss CAT and WAG models as if they are
exclusive, but these particular terms apply to non-exclusive model
components-- CAT refers to equilibrium frequency variation across sites
and WAG a particular exchangeability matrix. In this literature, CAT is
generally shorthand for Poisson+CAT and WAG is shorthand for
WAG+homogeneous equilibrium frequency estimation. To avoid confusion on
this point, we always specify the exchangeability matrix first, followed
by modifiers that describe the accommodation of heterogeneity in
equilibrium frequencies (e.g., CAT). Moreover, the Gamma-rate
heterogeneity is used in almost every analysis conducted here. If there
are no modifiers, then it is implied that site homogeneous models are
used. A detailed description of each component is provided in
Supplementary Results.

\hypertarget{gene-and-taxa-sampling}{%
\subsubsection{Gene and taxa sampling}\label{gene-and-taxa-sampling}}

High-throughput sequencing allows investigators interested in the
phylogeny of early diverging animals to assemble matrices with hundreds
or thousands of genes (Table 1). Surprisingly, gene overlap between
studies has rarely been assessed. Though there are many genes with
homologs present in most known animal genomes (\emph{e.g.}, animal BUSCO
genes\textsuperscript{4}), studies of animal phylogeny often relied on a
wide variety of different bioinformatic approaches to identifying and
selecting genes for their matrices. These take different approaches to
sequence comparison, filtering, and orthology assessment, and gene
selection can be impacted by which datasets are considered (see
supplementary text for more details). As a result, the genes selected
for analysis can vary widely between studies.

Outgroup selection (the non-animal taxa used to root the animal tree)
can affect the phylogenetic inference of animal evolution (Fig. 1C).
Several studies have progressively removed more distantly related
outgroup taxa to test the effect of outgroup selection to ingroup
topology\textsuperscript{5,6}. Three progressively more inclusive clades
have often been investigated: Choanomalia (animals plus most closely
related Choanoflagellata), Holoza (Choanimalia plus more distantly
related holozoans) and Opisthokonta (Holozoa + Fungi). It has been
suggested the inclusion of more distant related (e.g.~Fungi) outgroup
taxa tend to recover Ctenophora-sister, whereas increasing support of
Porifera-sister often recovered when more closely related outgroup taxa
were used with site-heterogeneous Poisson-CAT models\textsuperscript{6}.
In contrast of outgroup selection, sensitivity to ingroup sampling has
received less attention.

@@

Compositional heterogeneity (clade-specific and site-specific
differences in amino acid frequencies) have been a concern in the study
of phylogenetics, and has been of particular concern for the deep animal
phylogeny\textsuperscript{7}. Variation in evolutionary rates across
taxa can lead to compositional heterogeneity that may lead to long
branch attraction of sponges or ctenophores\textsuperscript{8}. Feuda
\emph{et al}. 2017\textsuperscript{9} has proposed to use data-recoding
method to reduce compositional heterogeneity in matrices. By discarding
information on which specific amino acid is found at each site in each
species and instead focusing on which groups those amino acids belong
to, they sought to reduce the impact of differences between species in
which particular amino acids within those groups are most frequent.
However, how the recoding methods impact phylogenetic inference has not
been well explored. Moreover, a recent simulation study has suggested
that non-recoding approaches significantly outperformed recoding
strategies\textsuperscript{10}.

@@

Here, we synthesize data and results from previous phylogenomic analyses
that tested Ctenophora-sister and Porifera-sister from 14 studies (Table
1). These matrices include all relevant studies published before 2020
for which we could obtain data matrices with gene partitions that
described boundaries between genes.

@@

Taxa sampling plays a critical role in phylogenetic inference,
especially for the taxa whose phylogenetic position is poorly
understood. However, several data matrices have relatively skewed taxon
sampling, with more than 50\% taxa sampled in Bilateria compared to
other major animal lineages (Extended Data Fig. 1). These skewed data
matrices include two studies constructed from ESTs
(Dunn2008\textsuperscript{2} and Hejnol2009\textsuperscript{11}) and one
study from whole-genome data (Borrowiec2015\textsuperscript{12}).
Importantly, several earlier studies before 2015 only include less than
five ctenophore species (Extended Data Fig. 1). The rest data matrices
have relatively balanced taxon composition, although the total number of
sampled taxon varies. In general, increasing taxon sampling is broadly
accepted to improve phylogenetic inference and aid in the placement of
the earliest diverging animal lineage. Thus, it should be noted that
Whelan2017\textsuperscript{13} (has the most Ctenophore taxa) and
Simion2017\textsuperscript{14} (contains the most taxon sampling) have
much broader taxon sampling than other matrices.

\includegraphics{manuscript_files/figure-latex/figure_taxon_rectangles-1.pdf}

\textbf{Fig. XXtaxon\_rectangles.} Matrix taxa and gene sampling. Each
of the primary matrices considered here, color coded by taxon sampling.
Horizontal size is proportional to the number of genes sampled, vertical
size to the number of taxa sampled.

\hypertarget{matrix-gene-sampling-are-largely-different-across-studies}{%
\subsubsection{Matrix gene sampling are largely different across
studies}\label{matrix-gene-sampling-are-largely-different-across-studies}}

We first used the Metazoa BUSCO genes\textsuperscript{4} to evaluate
gene composition from all data matrices and the result is unexpected
(Extended Data Fig. 2A). The animal BUSCO dataset contains 978
single-copy orthologs (identified from 64 animal genomes) that are most
likely shared by most animal genomes, and are commonly used to assess
the completeness of assemblies from genomic data\textsuperscript{4}. The
BUSCO genes also are frequently used as markers in many large-scale
phylogenomic studies (e.g., yeasts\textsuperscript{15},
spiders\textsuperscript{16}). Thus, it is reasonable to expect that
these conserved genes should present at a higher percentage than
non-BUSCO genes in most matrices. However, we found that gene partitions
from all matrices comprised less than 50\% of BUSCO genes, even with
several data matrices contain less than 20 percent (Extended Data Fig.
2B). Moreover, we identified number of ribosomal RNA genes in each data
matrix (Extended Data Fig. 2C) and found that Philippe2009,
Nosenko2013\_ribo\_14615 and Chang2015 matrices possessed much higher
content of ribosomal genes than other matrices (Extended Data Fig. 2D).
We also annotated each gene partition against
SwissProt\textsuperscript{{\textbf{???}}} and Gene Ontology databases,
and results are shown in Supplementary Table 1. Moreover, our network
analyses show that surprisingly little overlap of gene samplings across
all the data matrices from different studies (Fig. 2). Thus, these
results strongly suggested that gene sampling largely varies across
studies, indicating that different regions of the genomes are sampled
with little overlap among different data matrices.

@@

\includegraphics{manuscript_files/figure-latex/figure_support_analyses_published-1.pdf}
\textbf{Fig. XXfigure\_support\_analyses.} A summary of phylogenomic
results from the previous studies and reanalyses conducted in this
study. (A) A total of 135 analyses were transcribed from the literature
(Table XX). (B) A total of 102 new phylogenomic analyses were conducted
in this study (Supplementary Table XX).

Here we summarized all the 163 previous phylogenetic analyses from 15
studies (Fig. 3A and Supplementary Table 2). Among these studies, the
main conclusions of five studies strongly in favor of Porifera-sister
and 10 in support of Ctenophora-sister (Table 1). Moreover, three
studies are secondary studies without any new data matrices.
Interestingly, we found that all analyses strongly support of
Ctenophora-sister once site-homogeneous models were used, no matter what
outgroup choice, phylogenetic inference and whether data-recoding are
used. The support for Porifera-sister increases with the exclusion of
more distantly related outgroups and with the use of site-heterogeneious
CAT (Poisson+CAT or GTR+CAT) models in several matrices, including
Philippe2009, Ryan2013\_est and Whelan2015. It should be noted that many
BI analyses with CAT model are not well resolved (either not reach
convergence or do not provide significant support for the key node of
animal evolution) mainly due to computational limits of CAT models
employed in PhyloBayes. The strong support of Porifera-sister can only
be recovered from analyses using Poisson+CAT and recoding methods with
GTR+CAT models in several matrices. Although GTR+CAT model generally has
better model adequacy than Poisson+CAT or site-homogeneous models from
previous studies\textsuperscript{6,9}, Porifera-sister is not always
supported since several analyses strongly in favor Ctenophora-sister
even with only choanofalgellates are used as outgroups (Fig. 3A). Thus,
our summary of previous phylogenetic analyses strongly indicated that
the trend of the support of Porifera-sister is generally increasing with
the use of more complex CAT models and more closely related outgroups in
several representative matrices, whereas Ctenophora-sister is supported
in most other conditions.

\hypertarget{new-analyses-of-published-matrices-are-largely-consistent-with-previous}{%
\subsection{New analyses of published matrices are largely consistent
with
previous}\label{new-analyses-of-published-matrices-are-largely-consistent-with-previous}}

One of the challenges of interpreting support for the placement of the
animal root is that different programs, software versions, and settings
have been used across studies, and phylogenetic analysis decisions have
been approached in very different ways. Thus, it easily causes
confusions with the increase number of studies. Here we first reanalyze
the primary matrices from each study under consistent conditions with
IQtree\textsuperscript{17} under a panel of evolutionary models. We
selected this tool because it has greater model flexibility,
computational time and accuracy than other tools in maximum likelihood
(ML) framework\textsuperscript{18} (Fig. 3B; Supplementary Table 3).
Importantly, we also include a CAT-like, site-heterogeneous C10 -C60
(C60) models\textsuperscript{19} that are implemented in both ML
framework in our analyses and it is interesting to compare results
between C60 model to CAT models since it has never been used in any
matrices analyzed here.
